The Gillespie Algorithm: Stochastic Modeling of Molecular Reaction Networks

### 2024-11-28 paper/Gillespie-Algorithm.Rmd
Final version

In the following sections, we will first provide an overview of the Gillespie Algorithm and its historical development in the [Section 2](#sec-Topic-Review). Next, the theoretical underpinnings of the algorithm, including its basis in the master equation and stochastic modeling principles, will be detailed in the [Section 3](#sec-Theory). We will then demonstrate the application of the Gillespie Algorithm to two molecular systems in the [Section 4](#sec-application): the photodimerization of thymine, a DNA damage process induced by UV radiation (@MartinezFernandez2022), and the stochastic decay of radioactive isotopes (@Mouret2006). These simulations will be performed using the R programming language (@citeR). Finally, our [Section 5](#sec-conclusion) will summarize the findings, discuss the algorithm's broader implications, and explore potential future applications.

@@

```{r simulation_1, echo=FALSE, warning=FALSE, message=FALSE}
# Load required libraries
if (!require("reshape2")) install.packages("reshape2")
if (!require("ggplot2")) install.packages("ggplot2")
library(reshape2)
library(ggplot2)
# Define reaction rates
c1 <- (37 / 1e6) * 10             # Formation of CPD
c2 <- (2 / 1e6) * 10              # Formation of 64-PP
# Initial state
state <- c(T1 = 10000, T2 = 10000, CPD = 0, `64PP` = 0)
# Define propensity functions
propensities <- function(state) {
  c(
    c1 * state["T1"] * state["T2"], # Reaction 1
    c2 * state["T1"] * state["T2"]  # Reaction 2
  )
}
# Gillespie simulation function
simulate_gillespie <- function(state, max_time) {
  # Initialize time and output data frame
  time <- 0
  output <- data.frame(
    Time = numeric(0),
    T1 = numeric(0),
    T2 = numeric(0),
    CPD = numeric(0),
    `64PP` = numeric(0)
  )

@@

```{r simulation_2, echo=FALSE, warning=FALSE, message=FALSE}
# Define parameters
c <- 0.01  # Decay rate constant
state <- c(R = 10000)  # Initial radioactive substance
max_time <- 1000  # Extended maximum simulation time

>>>>

In the following sections, we will first provide an overview of the Gillespie Algorithm and its historical development in the [Section 2](#sec-Topic-Review). Next, the theoretical underpinnings of the algorithm, including its basis in the master equation and stochastic modeling principles, will be detailed in the [Section 3](#sec-Theory). We will then demonstrate the application of the Gillespie Algorithm to two molecular systems in the [Section 4](#sec-application): the photodimerization of thymine, and the stochastic decay of radioactive isotopes. These simulations will be performed using the R programming language (@citeR). Finally, our [Section 5](#sec-conclusion) will summarize the findings, discuss the algorithm's broader implications, and explore potential future applications.

@@

```{r simulation_1, echo=FALSE, warning=FALSE, message=FALSE}
# Environment set-up
library(reshape2)
library(ggplot2)
# Define reaction rates
c1 <- (37 / 1e6) * 10             # Formation of CPD
c2 <- (2 / 1e6) * 10              # Formation of 64-PP
# Initial state
state <- c(T1 = 10000, T2 = 10000, CPD = 0, `64PP` = 0)
# Define propensity functions
propensities <- function(state) {
  c(
    c1 * state["T1"] * state["T2"], # Reaction 1
    c2 * state["T1"] * state["T2"]  # Reaction 2
  )
}
# Gillespie simulation function
simulate_gillespie <- function(state, max_time) {
  # Initialize time and output data frame
  time <- 0
  output <- data.frame(
    Time = numeric(0),
    T1 = numeric(0),
    T2 = numeric(0),
    CPD = numeric(0),
    `64PP` = numeric(0)
  )

@@

```{r simulation_2, echo=FALSE, warning=FALSE, message=FALSE}
# Define parameters
c <- 0.01  # Decay rate constant
state <- c(R = 10000)  # Initial radioactive substance
max_time <- 1000
